---
title: "Beyond Basics: Elevate Your Plots"
subtitle: ""
author: "Dr. Thiyanga S. Talagala <br/>  Department of Statistics, Faculty of Applied Sciences <br/> University of Sri Jayewardenepura, Sri Lanka"
format:
  revealjs:
    width: 2000
    height: 1000
    margin: 0.1
    theme: [simple, custom.scss]
    transition: slide
    background-transition: fade
    slide-number: true
    show-slide-number: all
    title-slide-attributes: 
      data-background-color: "#800026"
      data-background-image: none
---

## Packages

```{r, message=FALSE, warning=FALSE, echo=TRUE}
library(ggplot2)
library(tibble)
# install.packages("devtools")
#devtools::install_github("thiyangt/elephants")
library(elephants)
#devtools::install_github("thiyangt/drone")
library(drone)
```

## Data

```{r, echo=TRUE}
elephants
```

```{r, echo=TRUE}
data("worldbankdata")
worldbankdata
```

# Pairwise plot matrix, scatterplot plot matrix, parallel coordinates plot, survival plot

## 

```{r}
#install.packages("GGally")
library(GGally)
ggpairs(elephants)
```

## 

```{r}
ggpairs(elephants, columns = c(2, 3, 4))
```

## 

```{r}
ggpairs(elephants, columns = c(2, 8, 9))
```

## 

```{r}
ggpairs(elephants, columns = c(2, 8, 9), mapping=aes(color=Gender))
```

## 

```{r}
ggpairs(elephants, columns = c(2, 8, 9), mapping=aes(color=Gender)) + scale_color_brewer(palette = "Dark2") 
```

## 

```{r}
ggpairs(elephants, columns = c(2, 8, 9), mapping=aes(color=Gender)) + scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2")
```

## 

```{r}
ggpairs(elephants, columns = c(2, 8, 9), mapping=aes(color=Gender)) + scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") + theme_bw()
```

## 

```{r}
ggpairs(elephants, columns = c(2, 8, 9), mapping=aes(color=Gender, alpha=0.5)) + scale_color_brewer(palette = "Dark2") + scale_fill_brewer(palette = "Dark2") + theme_bw()
```

---

# Combine separate ggplots into the same graphic

---

::: columns
::: {.column width="50%"}
```{r, echo=TRUE}
p1 <- ggplot(elephants, aes(x=Gender, fill=Gender)) + geom_bar()
p1
```
:::

::: {.column width="50%"}
```{r, echo=TRUE}
p2 <- ggplot(elephants, aes(x=Height, y=Fore_Feet_Circumference)) + geom_point() + coord_fixed(ratio=2)
p2
```
:::
:::

## 

```{r}
library(patchwork)
p1+p2
```

---

# Arranging plots into a grid

---


::: columns
::: {.column width="50%"}
```{r, echo=TRUE}
p1 <- ggplot(elephants, aes(x=Gender, fill=Gender)) + geom_bar()
p1
```
:::

::: {.column width="50%"}
```{r, echo=TRUE}
p2 <- ggplot(elephants, aes(x=Height, y=Fore_Feet_Circumference)) + geom_point() + coord_fixed(ratio=2)
p2
```
:::
:::

## 

```{r}
library(cowplot)
plot_grid(p1, p2, labels = c('A', 'B'), label_size = 12)
```

# Creating graphics with details from statistical tests

Defaults return

✅ raw data + distributions ✅ descriptive statistics ✅ inferential statistics ✅ effect size + CIs ✅ pairwise comparisons ✅ Bayesian hypothesis-testing ✅ Bayesian estimation

## 

```{r}
library(ggstatsplot)
ggbetweenstats(
  data  = elephants,
  x     = Gender,
  y     = Height,
  title = "Distribution of height by gender"
)
```

## 

To visualize the distribution of a single variable and check if its mean is significantly different from a specified value with a one-sample test

```{r}
gghistostats(
  data       = elephants,
  x          = Height,
  title      = "Height of Elephants",
  test.value = 2.5,
  binwidth   = 0.5
)
```

# Publication Ready Plots

## 

```{r}
library("ggpubr")
gghistogram(elephants, x = "Height",
   add = "mean", rug = TRUE,
   color = "Gender", fill = "Gender",
   palette = c( "#e7298a","#081d58"))
```


##

```{r}
ggboxplot(elephants, y = "Height", x="Gender",
   color = "Gender",
   palette = c( "#e7298a","#081d58"),
   add = "jitter", shape = "Gender")
```


#  Interactively explore your data by visualizing it with the ggplot2 package

## In-class demo

```r
library("esquisse")
esquisse::esquisser(elephants::elephants)
```

---

#  Add graphical information about one of the main panel’s axis.

---

```{r}
library(ggside)
ggplot(elephants, aes(x=Height, y=Fore_Feet_Circumference, colour = Gender)) + 
  geom_point(size = 2) +
  geom_xsideboxplot(aes(y =Gender), orientation = "y") +
  geom_ysidedensity(aes(x = after_stat(density)), position = "stack") +
  scale_ysidex_continuous(guide = guide_axis(angle = 90), minor_breaks = NULL) +
  theme(ggside.panel.scale = .3)
```

---

```{r}
library(ggside)
ggplot(elephants, aes(x=Height, y=Weight, colour = Gender)) + 
  geom_point(size = 2) +
  geom_xsideboxplot(aes(y =Gender), orientation = "y") +
  geom_ysidedensity(aes(x = after_stat(density)), position = "stack") +
  scale_ysidex_continuous(guide = guide_axis(angle = 90), minor_breaks = NULL) +
  theme(ggside.panel.scale = .3)
```

